\chapter*{Acknowledgements}

Firstly, I would like to express my sincere gratitude to my supervisor, Prof. Dr. Ivan G. Costa, for providing me with the opportunity to pursue my Ph.D. studies at the Institute for Computational Genomics at RWTH Aachen University. His unwavering support has been instrumental in the realization of this thesis.

I would also like to extend my appreciation to Prof. Dr. Thomas Berlage and Prof. Dr. xxxx for graciously agreeing to serve as the referees for the examination of my thesis. Special thanks to Prof. Dr. xxxx for serving as the chairperson during the examination procedure.

Moreover, I am grateful to several individuals whose contributions have played a significant role in this work. Prof. Dr. Rafael Kramann and Dr. Christoph Kuppe provided valuable advice and assistance throughout the project. I would like to acknowledge Jitske and others for their contributions.

Nadine Kaesler, David Schumacher, Eleni Stamellou...


A word of thanks goes to my colleagues who assisted with the analysis and proofreading of this study, as well as engaging in fruitful discussions: Dr. Zhijian Li, M.Sc. James Nagai, M.Sc. Martin Grasshoff, and M.Sc. Tiago Maie.


I am particularly appreciative of the collaboration with Tina Appel, whose assistance was invaluable during the preparation process for the defense.

My deepest thanks are reserved for my family and friends for their unwavering support during my time in Aachen. I am especially grateful to my parents, who provided me with the best possible start in life and supported me in every situation. Finally, heartfelt thanks to my wife xxx for her unwavering support throughout this journey.